How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
very small discussion update
Bass changes upto and including pag 3

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Research works involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively. Recent works has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. A recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common. Specifically, the research questions that we are trying to answer are:

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects. Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

@@

\section{Results}
\label{sec:results}
In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.

@@

In Table \ref{tbl-size} we summarize statistics for the analyzed metrics. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.

@@

\section{Discussion}
\label{sec:discussion}
\subsection{Implications}
We believe a large scale study of programs like ours can help language designers to tailor their language. In this section we highlight directions in which our study could support language design.

>>>>

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Previous work involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} has evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively. Recent works has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. A recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common. To address this goal, we answer the following research questions:

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} contains two one sprites, the cat, which is Scratch's default sprite and a Piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects. Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due to technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

@@

\section{Results}
\label{sec:results}
In the following sections, for each of the research questions, we describe the results obtained through the analysis of the  \nAnalyzedPrograms~Scratch projects in our dataset.

@@

In Table \ref{tbl-size} we summarize statistics for the analyzed metrics. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts} and \ref{fig:size-locs} plot the distribution of these size metrics.

@@

\section{Discussion}
\label{sec:discussion}
\subsection{Implications}
We believe a large scale study of programs like ours can help language designers to tailor their language. In this section we highlight directions in which our study could support language design. There are many other implications to be considered, which is why we have made our dataset public.
